---
title: "Lidar Birdies"
author: "Mike Allen"
date: "11/10/2020"
output: html_document
---

```{r}
library(raster)
library(rgdal)
library(sp)
library(sf)
library(tidyverse)
library(lidR)
library(tmap)
library(tmaptools)
```
# Read in shapefiles
```{r}
# read in duke field borders
duke <- 
   read_sf("data/duke_GL_pred_areas.shp") %>%
   st_transform(crs = 32618)
```
# Read in and check 2008 lidar files
Note: reprojected into UTM zone 18 in LAStools in QGIS.
source for code: http://www.wvview.org/spatial_analytics/lidar/_site/index.html#lidar
```{r}
las_cat <- readLAScatalog("data/lidar2008_utm/")
projection(las_cat) <- crs(32618)
summary(las_cat)

lascheck(las_cat)
```
# Choose scale to process 2008 lidar files
```{r}
opt_chunk_size(las_cat) <- 2000
plot(las_cat, chunk_pattern = TRUE)

opt_chunk_buffer(las_cat) <- 40
plot(las_cat, chunk_pattern = TRUE)
```
# View summary of 2008 lidar files
```{r}
summary(las_cat)
```
# Create & view 2008 digital terrain models ("bare earth")
```{r}
# opt_output_files(las_cat) <- "data/dtm2008/dtm_{XLEFT}_{YBOTTOM}"
# dtm2008 <-
#    grid_terrain(las_cat,
#                 res = 1,
#                 knnidw(k = 10, p = 2),
#                 keep_lowest = FALSE) %>%
#    crop(y = (extent(duke) + c(-250, 250, -250, 250)))
# crs(dtm2008) <- crs(duke)

dtm2008 <- 
  raster("data/dtm2008.tif") # to save time

slope <- terrain(dtm2008, opt='slope')
aspect <- terrain(dtm2008, opt='aspect')
hs <- hillShade(slope, aspect, angle=45, direction=315)

tm_shape(hs)+
tm_raster(style= "cont", palette=get_brewer_pal("Greys", plot=FALSE))+
tm_layout(legend.outside = TRUE)
```
# Create 2008 digital surface models ("bare earth + canopy height")
```{r}
# opt_output_files(las_cat) <- "data/dsm2008/dsm_{XLEFT}_{YBOTTOM}"
# dsm2008 <-
#    grid_canopy(las_cat, res = 1, pitfree(c(0, 2, 5, 10, 15), c(0, 1))) %>%
#    crop(extent(duke) + c(-250, 250, -250, 250))
# crs(dsm2008) <- crs(duke)

dsm2008 <- 
   raster("data/dsm2008.tif") # to save time

slope <- terrain(dsm2008, opt='slope')
aspect <- terrain(dsm2008, opt='aspect')
hs <- hillShade(slope, aspect, angle=45, direction=315)

tm_shape(hs)+
tm_raster(style= "cont", palette=get_brewer_pal("Greys", plot=FALSE))+
tm_layout(legend.outside = TRUE)
```
```{r}
dtm2008_plus_eye <- dtm2008 + 1.61 # the height of Mike's eyes

dsm.eye <- max(dtm2008_plus_eye, dsm2008)

# next: 
#   5) run r.horizon every 5 degrees in GRASS GUI 
#   6) make raster of max & mean (use r.series in GRASS GUI)
#   7)  test as covariate for psi
```

# Map mean horizon heights from all angles
Note: I used GRASS for this via the program's user interface.
```{r}
horizon_mean <- raster("data/horizon/horizon_10_mean.tif") %>%
   mask(duke)
plot(horizon_mean, col = viridis::inferno(256), zlim = c(0,30))
```






# To change raster regions manually, e.g., to remove tree lines
```{r}

boxes = rbind.data.frame(
expand.grid(y = c(7776:7806), x = c(8480:8500), new = 31.6),
expand.grid(y = c(7807:7844), x = c(8483:8503), new = 32.3),
expand.grid(y = c(7850:7902), x = c(8492:8512), new = 33.5),
expand.grid(y = c(7910:7980), x = c(8495:8530), new = 34.7),
expand.grid(y = c(7980:8033), x = c(8512:8642), new = 34.6),
expand.grid(y = c(7988:8022), x = c(8642:8712), new = 34.1),
expand.grid(y = c(8033:8119), x = c(8515:8549), new = 33.5),
expand.grid(y = c(8122:8164), x = c(8530:8554), new = 32.2)
) %>%
   group_by(y, x) %>%
   summarise(new = mean(new)) %>%
   ungroup() %>%
   mutate(y = format(y + 4480000.5, digits = 8), 
          x = x + 520000.5,
          keyx = as.character(x),
          keyy = as.character(y))

dp = rasterToPoints(dsm) %>%
   data.frame() %>%
   mutate(y = formatC(y, digits = 1, format = "f"),
          keyx = as.character(x),
          keyy = as.character(y)) %>%
   left_join(fix, by = c("x", "y", "keyx", "keyy")) %>%
   mutate(value.final = case_when(is.na(new) ~ final_duke_dsm,
                                  TRUE ~ new)) %>%
   dplyr::select(x, y, keyx, keyy, value.final) %>%
   left_join(boxes, by = c("x", "y", "keyx", "keyy")) %>%
   mutate(value.final2 = case_when(is.na(new) ~ value.final,
                                  TRUE ~ new)) %>%
   dplyr::select(x, y, value.final2) %>%
   mutate(y = as.numeric(y))

# get values
vals <- 
   dp$value.final2

# make the new raster with treelines and artifacts removed
dsm.fix <- dsm %>%
   setValues(vals)

#writeRaster(dsm.fix, "data/duke_dsm_fixed.tif")
```


