---
title:  "Code from 'A lidar-based openness index to aid conservation planning for grassland wildlife' (openness mapping)"
author: "Mike Allen"
date: "11/10/2021"
output: html_document
---
# Load packages and shapefile of Duke Farms fields 
R version 4.1.0 (2021-05-18)
Platform: x86_64-w64-mingw32/x64 (64-bit)
Running under: Windows 10 x64 (build 19043)
```{r}
library(raster) # raster_3.4-13 
library(rgdal) # rgdal_1.5-23 
library(sp) # sp_1.4-5 
library(sf) # sf_1.0-1   
library(tidyverse) # tidyverse_1.3.1
library(lidR) # lidR_3.1.4
library(tmap) # tmap_3.3-2
library(tmaptools) # tmaptools_3.1-1

# read in the Duke Farms field borders shapefile and transform to UTM
duke <- 
   read_sf("data/duke_GL_pred_areas.shp") %>%
   st_transform(crs = 32618) 

# create sf object of Duke Farms survey points
duke_pts <- read.csv("data/grsp.occupancy.openness.csv")[,1:4] %>% 
   dplyr::select(pt, latitude, longitude) %>%
   st_as_sf(coords = c("longitude", "latitude"),
            crs = 4326) %>%
   st_transform(crs = crs(duke))

# load template raster to project to (for converting to UTM)
template <- raster("data/template.tif")
```
# Read in and check the lidar files
The 9 lidar tiles used included: G8C4.laz, G8B13.laz, G8D1.laz, G8A16.laz, G8C3.laz, G8A15.laz, G8C8.laz, G8D5.laz, G8C7.laz. They were downloaded from the "Northwest New Jersey 2018" collection of New Jersey Geographic Information Network (see References). The 9 .laz files are about 750 MB total.
```{r}
# Read 9 lidar tiles from the folder lidar2018
las_cat18 <- readLAScatalog("data/lidar2018/")

# define their projection
projection(las_cat18) <- crs(6527) # state plane nad83(2011) in US feet

#print a summary of the files
summary(las_cat18)

# perform a check on the lidar data quality
lascheck(las_cat18)
```
# Choose scale to process the lidar files
Choose a 'chunk size' and overlap value to process point cloud data. Per the LiDAR and Image Processing in R" (see References): "It is generally good to define an overlap to avoid any gaps in generated raster grid surfaces. You will need to experiment with appropriate chunk sizes. This will depend on the point density of the data and the available memory in your computer. The goal is to provide small enough tiles so that you will not run out of memory but not too small to produce a lot of tiles and process the data inefficiently."
```{r}
# choose a 'chunk size' for processing and plot what it looks like
# note: this is in units of US feet for my files
opt_chunk_size(las_cat18) <- 5100 
plot(las_cat18, chunk_pattern = TRUE)

# choose an overlap value for the chunks for processing and plot what it looks like
opt_chunk_buffer(las_cat18) <- 100
plot(las_cat18, chunk_pattern = TRUE)
```
# View a summary of the lidar catalog again
To confirm chunk size and overlap settings.
```{r}
 summary(las_cat18)
```
# Create & view the "bare earth" digital terrain models (DTM) from the point cloud
```{r}
# create directory for DTM intermediate files
if(dir.exists("data/dtm2018/")==F){dir.create("data/dtm2018/")}

# define the output folder and file naming convention
opt_output_files(las_cat18) <- "data/dtm2018/dtm_{XLEFT}_{YBOTTOM}"

# make a DTM with resolution of 1 meter
dtm2018_step1 <-
   grid_terrain(las_cat18,
                res = 3.28084,
                # this converts feet to meters to make 1 m res
                knnidw(k = 10, p = 2),
                keep_lowest = FALSE)

# define projection
crs(dtm2018_step1) <-
   CRS("+init=epsg:6527") # nj state plane nad83 (2011), grs80 in US feet

# reproject the raster to UTM
# used a premade UTM raster cropped to Duke for this purpose
dtm2018 <-
   projectRaster(from = dtm2018_step1,
                 to = template)

# convert vertical units from feet to meters
dtm2018 <- dtm2018 / 3.28084 

# assign name to DTM
names(dtm2018) <- "dtm2018"
```
# View the DTM as a hillshade
```{r}
slope <- terrain(dtm2018, opt='slope')
aspect <- terrain(dtm2018, opt='aspect')
hs <- hillShade(slope, aspect, angle=45, direction=315)

tm_shape(hs)+
tm_raster(style= "cont", palette=get_brewer_pal("Greys", plot=FALSE))+
tm_layout(legend.outside = TRUE)
```
# Create digital surface models, DSM (i.e., "bare earth + canopy height")
```{r}
# create directory for DSM intermediate files
if(dir.exists("data/dsm2018/")==F){dir.create("data/dsm2018/")}

# set output folder and naming convention for DSM
opt_output_files(las_cat18) <-
   "data/dsm2018/dsm_{XLEFT}_{YBOTTOM}"

# calculate DSM
dsm2018_step1 <-
   grid_canopy(las_cat18, res = 3.28084, pitfree(c(0, 2, 5, 10, 15), c(0, 1)))

# remove elevation outliers
dsm2018_step1[dsm2018_step1>300] <- NA 

# define projection of DSM
crs(dsm2018_step1) <-
    CRS("+init=epsg:6527") # nj state plane nad83 (2011), grs80 in US feet

# project to UTM using template raster
 dsm2018 <-
    projectRaster(from = dsm2018_step1,
                  to = template)

 # convert vertical units to meters
 dsm2018 <- dsm2018 / 3.28084
 
 # fix name of DSM
 names(dsm2018) <- "dsm2018"
```
# View the DSM as a hillshade
```{r}
slope <- terrain(dsm2018, opt='slope')
aspect <- terrain(dsm2018, opt='aspect')
hs <- hillShade(slope, aspect, angle=45, direction=315)

tm_shape(hs)+
tm_raster(style= "cont", palette=get_brewer_pal("Greys", plot=FALSE))+
tm_layout(legend.outside = TRUE)
```
# Remove a tree line that was not present during the bird surveys
```{r}
# remove a tree line that was removed in 2018 after lidar but before survey
# make 4 boxes to "cut down" the tree line
boxes2018 = rbind.data.frame(
   expand.grid(
      y = c(7470:7610),
      x = c(8635:8681),
      new = 30.5
   ),
   expand.grid(
      y = c(7408:7465),
      x = c(8627:8661),
      new = 29.5
   ),
   expand.grid(
      y = c(7352:7410),
      x = c(8619:8649),
      new = 27.5
   ),
   expand.grid(
      y = c(7302:7354),
      x = c(8616:8646),
      new = 25.5
   )
) %>%
   group_by(y, x) %>%
   summarise(new = mean(new)) %>%
   ungroup() %>%
   mutate(y = format(y + 4480000.5, digits = 8),
          x = x + 520000.5)

dsm2018.newVals_step1 <- dsm2018
dsm2018.newVals_step1[is.na(dsm2018.newVals_step1)] <- 0
dsm2018.newVals <- rasterToPoints(dsm2018.newVals_step1) %>%
   as.data.frame(headers = F) %>%
   mutate(y = as.character(y)) %>%
   left_join(boxes2018, by = c("x", "y")) %>%
   mutate(dsm2018 = case_when(is.na(new) == T ~ dsm2018,
                              TRUE ~ new),
          y = as.numeric(y)) %>%
   dplyr::select(-new)

# make the new raster with treelines and artifacts removed
dsm2018.fix <- dsm2018 %>%
   setValues(dsm2018.newVals[, 3])
```
# Remove powerlines to test if occupancy model fits better with or without this
If without, then they are not likely reacting strongly to the wires
```{r}
boxes2018.nowires = rbind.data.frame(
   expand.grid(
      y = c(8799:8929),
      x = c(9301:9358),
      new = 3
   ),
   expand.grid(
      y = c(8619:8800),
      x = c(9292:9332),
      new = 3
   ),
   expand.grid(
      y = c(8438:8623),
      x = c(9282:9315),
      new = 3 
   ),
   expand.grid(
      y = c(8260:8442),
      x = c(9272:9309),
      new = 3
   ),
   expand.grid(
      y = c(8130:8263),
      x = c(9264:9296),
      new = 3
   ),
   expand.grid(
      y = c(7987:8132),
      x = c(9257:9286),
      new = 3
   ),
   expand.grid(
      y = c(7853:7992),
      x = c(9249:9282),
      new = 3 
   ),   
   expand.grid(
      y = c(7702:7861),
      x = c(9242:9314),
      new = 3 
   ),
   expand.grid(
      y = c(7684:7707),
      x = c(9276:9336),
      new = 3
   )
) %>%
   group_by(y, x) %>%
   summarise(new = mean(new)) %>%
   ungroup() %>%
   mutate(y = format(y + 4480000.5, digits = 8),
          x = x + 520000.5)

dsm2018.nowires.newVals <- rasterToPoints(dsm2018.fix) %>%
   as.data.frame(headers = F) %>%
   mutate(y = as.character(y)) %>%
   left_join(boxes2018.nowires, by = c("x", "y")) %>%
   mutate(dsm2018 = case_when(is.na(new) == T ~ dsm2018,
                              TRUE ~ new),
          y = as.numeric(y)) %>%
   dplyr::select(-new)

# make the new raster with treelines and artifacts removed
dsm2018.nowires <- dsm2018.fix %>%
   setValues(dsm2018.nowires.newVals[,3])
```
# Remove all tree lines for scenario evaluation
To evaluate effect on predicted occupancy
```{r}
boxes2018.notrees = rbind.data.frame(
   expand.grid( # SW tree line
      y = c(7320:7676),
      x = c(8483:8677),
      new = 3
   ),
   expand.grid( # SE tree line
      y = c(7545:7682),
      x = c(8698:8853),
      new = 3
   ),
   expand.grid( # N tree line
      y = c(8356:8702),
      x = c(8642:8805),
      new = 3
   )
) %>%
   group_by(y, x) %>%
   summarise(new = mean(new)) %>%
   ungroup() %>%
   mutate(y = format(y + 4480000.5, digits = 8),
          x = x + 520000.5)

dsm2018.notrees.newVals <- rasterToPoints(dsm2018.fix) %>%
   as.data.frame(headers = F) %>%
   mutate(y = as.character(y)) %>%
   left_join(boxes2018.notrees, by = c("x", "y")) %>%
   mutate(dsm2018 = case_when(is.na(new) == T ~ dsm2018,
                              TRUE ~ new),
          y = as.numeric(y)) %>%
   dplyr::select(-new)

# make the new raster with treelines and artifacts removed
dsm2018.notrees <- dsm2018.fix %>%
   setValues(dsm2018.notrees.newVals[,3])
```
# Remove just the SW tree line for scenario evaluation
To evaluate effect on predicted occupancy
```{r}
boxes2018.noSW = rbind.data.frame(
   expand.grid( # SW tree line
      y = c(7320:7676),
      x = c(8483:8677),
      new = 3
   )
) %>%
   group_by(y, x) %>%
   summarise(new = mean(new)) %>%
   ungroup() %>%
   mutate(y = format(y + 4480000.5, digits = 8),
          x = x + 520000.5)

dsm2018.noSW.newVals <- rasterToPoints(dsm2018.fix) %>%
   as.data.frame(headers = F) %>%
   mutate(y = as.character(y)) %>%
   left_join(boxes2018.noSW, by = c("x", "y")) %>%
   mutate(dsm2018 = case_when(is.na(new) == T ~ dsm2018,
                              TRUE ~ new),
          y = as.numeric(y)) %>%
   dplyr::select(-new)

# make the new raster with treelines and artifacts removed
dsm2018.noSW <- dsm2018.fix %>%
   setValues(dsm2018.noSW.newVals[,3])
```
# Remove just the SE tree line for scenario evaluation
To evaluate effect on predicted occupancy
```{r}
boxes2018.noSE = rbind.data.frame(
   expand.grid( # SE tree line
      y = c(7545:7682),
      x = c(8698:8853),
      new = 3
   )
) %>%
   group_by(y, x) %>%
   summarise(new = mean(new)) %>%
   ungroup() %>%
   mutate(y = format(y + 4480000.5, digits = 8),
          x = x + 520000.5)

dsm2018.noSE.newVals <- rasterToPoints(dsm2018.fix) %>%
   as.data.frame(headers = F) %>%
   mutate(y = as.character(y)) %>%
   left_join(boxes2018.noSE, by = c("x", "y")) %>%
   mutate(dsm2018 = case_when(is.na(new) == T ~ dsm2018,
                              TRUE ~ new),
          y = as.numeric(y)) %>%
   dplyr::select(-new)

# make the new raster with treelines and artifacts removed
dsm2018.noSE <- dsm2018.fix %>%
   setValues(dsm2018.noSE.newVals[,3])
```
# Remove just the N tree line for scenario evaluation
To evaluate effect on predicted occupancy
```{r}
boxes2018.noN = rbind.data.frame(
   expand.grid( # N tree line
      y = c(8356:8702),
      x = c(8642:8805),
      new = 3
   )
) %>%
   group_by(y, x) %>%
   summarise(new = mean(new)) %>%
   ungroup() %>%
   mutate(y = format(y + 4480000.5, digits = 8),
          x = x + 520000.5)

dsm2018.noN.newVals <- rasterToPoints(dsm2018.fix) %>%
   as.data.frame(headers = F) %>%
   mutate(y = as.character(y)) %>%
   left_join(boxes2018.noN, by = c("x", "y")) %>%
   mutate(dsm2018 = case_when(is.na(new) == T ~ dsm2018,
                              TRUE ~ new),
          y = as.numeric(y)) %>%
   dplyr::select(-new)

# make the new raster with treelines and artifacts removed
dsm2018.noN <- dsm2018.fix %>%
   setValues(dsm2018.noN.newVals[,3])
```
# Make the final rasters for horizon analysis
These are the corrected DSMs "flooded" to the height of the DTM plus 1.61 meters (human eye height).
```{r}
# add 1.61 m to every cell in the DTM
dtm2018_plus_eye <- dtm2018 + 1.61 # eye height

# 'flood' each DSM to 1.61 m from ground height for horizon analysis
dsm.eye2018.fix <- max(dtm2018_plus_eye, dsm2018.fix)
dsm.eye2018.nowires <- max(dtm2018_plus_eye, dsm2018.nowires)
dsm.eye2018.notrees <- max(dtm2018_plus_eye, dsm2018.notrees)
dsm.eye2018.noSW <- max(dtm2018_plus_eye, dsm2018.noSW)
dsm.eye2018.noSE <- max(dtm2018_plus_eye, dsm2018.noSE)
dsm.eye2018.noN <- max(dtm2018_plus_eye, dsm2018.noN)

# write these to tif files for horizon analysis in GRASS GIS
#writeRaster(dsm.eye2018.fix, "data/dsm.eye2018.fix.tif", overwrite = T)
#writeRaster(dsm.eye2018.nowires, "data/dsm.eye2018.nowires.tif", overwrite = T)
#writeRaster(dsm.eye2018.notrees, "data/dsm.eye2018.notrees.tif", overwrite = T)
#writeRaster(dsm.eye2018.noSW, "data/dsm.eye2018.noSW.tif", overwrite = T)
#writeRaster(dsm.eye2018.noSE, "data/dsm.eye2018.noSE.tif", overwrite = T)
#writeRaster(dsm.eye2018.noN, "data/dsm.eye2018.noN.tif", overwrite = T)
```
# Map mean horizon heights from all angles
Note: In this step, I used the GRASS GIS GUI (r.horizon & r.series functions) to make the horizon rasters. Mean or max of 72 height-of-horizon measurements around each cell.

# Load horizon rasters created using the GRASS GIS GUI 
Rasters were created based on the "dsm.eye" rasters as described above
```{r}
horizon2018_mean <- raster("data/horizon_rasters/horizon2018fix_mean_5.tif") 
horizon2018_max <- raster("data/horizon_rasters/horizon2018fix_max_5.tif") 

plot((90-mask(horizon2018_mean, duke)), col = viridis::inferno(256))
plot((90-mask(horizon2018_max, duke)), col = viridis::inferno(256))
```
# Formatting openness rasters for site-level covariate in occupancy analysis
```{r}
# function to create raster of circular moving window average of openness
mean.open = function(raster, 
                       radius) {
  # create the circular buffers (as matrices for the focal function)
  circles <- focalWeight(raster, radius, type = 'circle')
  circles[circles > 0] <- 1
  
  #raster[is.na(raster[])] <- 0
  
  mean.open.map = focal(
    raster,
    w = circles,
    fun = mean,
    pad = F,
    NAonly = F,
    na.rm = T
  )
  
  return(mean.open.map)
}

# load mean-angle horizon raster
horizon2018_mean <-
   90 - raster("data/horizon_rasters/horizon2018fix_mean_5.tif")

# create 40 m moving window average version
horizon2018_mean_cov <-
    mean.open(mask(horizon2018_mean, duke), radius = 40) %>%
   mask(duke)

# load max-angle horizon raster
horizon2018_max <-
   90 - raster("data/horizon_rasters/horizon2018fix_max_5.tif")

# create 40 m moving window average version
horizon2018_max_cov <-
   mean.open(mask(horizon2018_max, duke), radius = 40) %>%
   mask(duke)

# load mean-angle horizon raster with all trees removed
horizon2018_No_trees_mean <-
   90 - raster("data/horizon_rasters/horizon2018_No_trees_mean.tif")

# create 40 m moving window average version
horizon2018_No_trees_mean_cov <-
   mean.open(mask(horizon2018_No_trees_mean, duke), radius = 40) %>%
   mask(duke)

# load max-angle horizon raster with all trees removed
horizon2018_No_trees_max <-
   90 - raster("data/horizon_rasters/horizon2018_No_trees_max.tif")

# create 40 m moving window average version
horizon2018_No_trees_max_cov <-
   mean.open(mask(horizon2018_No_trees_max, duke), radius = 40) %>%
   mask(duke)

# load mean-angle horizon raster with powerlines removed
horizon2018_No_wires_mean <-
  90 - raster("data/horizon_rasters/horizon2018_No_wires_mean.tif")

# create 40 m moving window average version
horizon2018_No_wires_mean_cov <-
  mean.open(mask(horizon2018_No_wires_mean, duke), radius = 40) %>%
   mask(duke)

# load max-angle horizon raster with powerlines removed
horizon2018_No_wires_max <-
  90 - raster("data/horizon_rasters/horizon2018_No_wires_max.tif")

# create 40 m moving window average version
horizon2018_No_wires_max_cov <-
  mean.open(mask(horizon2018_No_wires_max, duke), radius = 40) %>%
   mask(duke)

# load mean-angle horizon raster with SW treeline removed
horizon2018_No_SW_mean <-
  90 - raster("data/horizon_rasters/horizon2018_No_SW_mean.tif")

# create 40 m moving window average version
horizon2018_No_SW_mean_cov <-
  mean.open(mask(horizon2018_No_SW_mean, duke), radius = 40) %>%
   mask(duke)

# load max-angle horizon raster with SW treeline removed
horizon2018_No_SW_max <-
  90 - raster("data/horizon_rasters/horizon2018_No_SW_max.tif")

# create 40 m moving window average version
horizon2018_No_SW_max_cov <-
  mean.open(mask(horizon2018_No_SW_max, duke), radius = 40) %>%
   mask(duke)

# load mean-angle horizon raster with SE treeline removed
horizon2018_No_SE_mean <-
  90 - raster("data/horizon_rasters/horizon2018_No_SE_mean.tif")

# create 40 m moving window average version
horizon2018_No_SE_mean_cov <-
  mean.open(mask(horizon2018_No_SE_mean, duke), radius = 40) %>%
   mask(duke)

# load max-angle horizon raster with SE treeline removed
horizon2018_No_SE_max <-
  90 - raster("data/horizon_rasters/horizon2018_No_SE_max.tif")

# create 40 m moving window average version
horizon2018_No_SE_max_cov <-
  mean.open(mask(horizon2018_No_SE_max, duke), radius = 40) %>%
   mask(duke)

# load mean-angle horizon raster with N treeline removed
horizon2018_No_N_mean <-
  90 - raster("data/horizon_rasters/horizon2018_No_N3_mean.tif")

# create 40 m moving window average version
horizon2018_No_N_mean_cov <-
  mean.open(mask(horizon2018_No_N_mean, duke), radius = 40) %>%
   mask(duke)

# load max-angle horizon raster with N treeline removed
horizon2018_No_N_max <-
  90 - raster("data/horizon_rasters/horizon2018_No_N3_max.tif")

# create 40 m moving window average version
horizon2018_No_N_max_cov <-
  mean.open(mask(horizon2018_No_N_max, duke), radius = 40) %>%
   mask(duke)
```
# Extract openness covariate values at survey points
```{r}
site.cov <- cbind.data.frame(pt = as.data.frame(duke_pts)$pt, 
           mean2018 = raster::extract(horizon2018_mean_cov, duke_pts),
           max2018 = raster::extract(horizon2018_max_cov, duke_pts),
           No_trees_mean = raster::extract(horizon2018_No_trees_mean_cov, duke_pts),
           No_trees_max = raster::extract(horizon2018_No_trees_max_cov, duke_pts),
           No_wires_mean = raster::extract(horizon2018_No_wires_mean_cov, duke_pts),
           No_wires_max = raster::extract(horizon2018_No_wires_max_cov, duke_pts),
           No_SW_mean = raster::extract(horizon2018_No_SW_mean_cov, duke_pts),
           No_SW_max = raster::extract(horizon2018_No_SW_max_cov, duke_pts),
           No_SE_mean = raster::extract(horizon2018_No_SE_mean_cov, duke_pts),
           No_SE_max = raster::extract(horizon2018_No_SE_max_cov, duke_pts),
           No_N_mean = raster::extract(horizon2018_No_N_mean_cov, duke_pts),
           No_N_max = raster::extract(horizon2018_No_N_max_cov, duke_pts)) 
```
# References

NJGIN websites for downloading "Northwest New Jersey 2018" lidar tiles
https://njgin.nj.gov/njgin/edata/elevation/index.html#!/

https://njogis-newjersey.opendata.arcgis.com/datasets/newjersey::northwest-nj-2018-lidar-tile-grid-hosted/explore?location=40.850139%2C-74.644373%2C9.19

Source for lidR code
http://www.wvview.org/spatial_analytics/lidar/_site/index.html#lidar

```

